---
title:Randomized-Controlled Treatments and Factor Designs.
output: html_document
---

```{r read-in-data}
library(readr)
natlterr2015_cleaned <- read_csv("natlterr2015_cleaned.csv")

#TODO Karla: make the data dictionary into a markdown table. I've added a dictionary from another file below for you to revise and then you can delete this.
# column name meaning:
#	birthweight (continuous in grams) - dbwt
# gestational age (in completed weeks) -combgest
#	Sex -sex
# birth month -dob_mm
# Smoker- CIG_REC 
```
### Description of the data

TODO Karla: Add a few sentences describing the data. (These are births from the US territories which include ...... from the year XXXX. These data were downloaded from [here](URL linked to the page). 

Here is the data dictionary for this dataset:

| Variable        | Description                                         |
|-----------------|-----------------------------------------------------|
| Country_Name    | The name of the country                             | 
| CountryCode     | 3-digit unique country identifier                   |                
| Births_Per_1000 | The number of births per 1000 people in the country |
| IncomeGroup     | The income group the country is categorized into    |
| Region          | The geographic region the country belongs to        |
| GDP.2006        | The gross domestic product (GDP) of the country in 2006 |
| CS.rate         | The percent of births that were delivered by cesarean section |

### Simulating a randomized controlled trial in newborns

TODO Corinne: to finish writing this: The purpose of this assignment is to take a **sample** from the **population** of newborns as if we were conducting a randomized controlled trial (RCT). Suppose that the US territories are interested in vaccinating their newborns against respiratory syncytial virus (RSV), which can cause severe lower respiratory infections in newborns and lead to hospitalizations and adverse respiratory outcomes. To conduct an RCT 

### Get familiar with the dataset

TODO: Have the students use str(), names(), head(), dim() to look at the data. Ask how many individuals in the data set there are, and who these individuals represent.

1. We will make a new column "treatment" and randomly assign patients to 0 (placebo) or 1 (active treatment). There are two ways to do this. 

* Way 1:
      * The first create a function that assigns 1 or 0 with probability 50% each and then applying it to your new column. I will help you get started with the function. We will be using the function sample(). The first input it takes is a vector with 1+ elements to choose from. In this case, it is only 0 and 1. The second argument we will input is the number of items we want to choose from. In this case, it will be 1, because we want 0 or 1, but not both. 
```{r}
# Do not forget to install the package "dplyr" using install.packages() before you try to load the library. 
random_sampler<-function(x) sample(c(0,1),1)
```

Your turn! Now apply this function to your column. You can do this by using the apply() function. Write ?apply to learn more. 
```{r}
#### YOUR CODE GOES HERE ####
```


* Way 2:
      *The second is to use the function rbinom() which takes three inputs. The first input is the length of the column. You can do this by checking the how many rows are in your dataframe using the function nrow(). The second is the number of trials, in this case, 1.The third is the probability, which we want it to be 0.5. Now you can assign this function to the new column in your data set. You can create the new column using the "df$colname", or "df[,'colname']
```{r}
#### YOUR CODE GOES HERE ####
```

Note: By the law of large numbers, assuming the data is large enough, the coin flip function (heads=1 | tails=0) will roughly sort half of the babies into treatment and half into the placebo group. The law says that if take a sample size n, from a random variable X, and we were to take the average of these samples, the sample mean will *approach* the expected value of the random variable (population mean) as n->inf. 
In this case, if Pr(X=1)=p=0.5, and Pr(X=0)=q=0.5 then E(X)=Pr(X=1)*1+Pr(X=0)*0=0.5


2.a Now produce a "Table 1", showing the descriptive statistics for treatment vs placebo. You can use the function table() on the treatment column to get frequency count. 
```{r}
#### YOUR CODE GOES HERE ####
```

2.b Comment on what you notice (the two results should be very close)
```{r}
#### YOUR COMMENTS GO HERE ####
```


3. Now, re-run the code you created in part 1. Use the function table() to get a frequency count for placebo and treatment, but do NOT save it to the variable called 'Table1'. You should be able to notice that the babies that were assigned to placebo or treatment in the first round were not necessarily assigned to the categories in the same round. You should also notice that the frequencies, although not equal are close to each other. Now try to convince yourself that you understand how the law of large numbers works when the data is large enough by noting how similar the descriptive statistics are on each run. 


4.a Now we will take a small sample from the pop (less than 100 observations), and randomly assign each to a treatment and placebo. I will get you started on taking the sample. We will use the function sample() and set.seed() 

```{r}
set.seed(123) #pseudo-random number generation. It ensured you can reproduce results
sample_index <- sample(1:nrow(natlterr2015_cleaned), 99) # pick any number < 100 
sample_index
natlterr2015_sample<-natlterr2015_cleaned[sample_index, ] # natlterr2015_sample is a random subset
```

4.b Your Turn! Run the random_sampler function on the subset data frame created above to get a fresh random assignment of placebo and treatment in the treatment column. 
```{r}
#### YOUR CODE GOES HERE ####
```

4.c Remake the "Table 1" with natlterr2015_sample results. You can run 4.b and 4.c multiple times to observe different results. 
```{r}
#### YOUR CODE GOES HERE ####
```

4.d Comment on what you notice
```{r}
#### YOUR COMMENTS GO HERE ####
```


5. Suppose that 17% of the babies were born to mothers who smoked during pregnancy. Under randomization of the original large dataframe, what percent of babies assigned to treatment were born to mothers who smoked during pregnancy? Under placebo?
```{r}
#### YOUR CODE GOES HERE ####
```
