---
title: "Workshop: Section 1 & Section 2"
output: html_document
---

### Notes:
* Might be useful to download the packages "ggplot2", "plotly" and"dplyr" before running any of R chuncks. You can use the function "install.packages()".  



#SECTION 1: 

If you haven't already, please follow the instructions in setup.md before proceeding!

Read in the first dataset
```{r}
BlackWhite_results <- read.csv("./BlackWhite_results.csv")
```

Let's first use the viewer to look at these data.
1. Click the data icon next to BlackWhite_results in the environment tab
Or run the code:
```{r}
View(BlackWhite_results)
```

These are some results from a study I did on the inequality in life expectancy between blacks and whites in the US.
Let's first examine the structure of the dataset:
```{r}
str(BlackWhite_results)
```

What are the first three variables?
Use the dollar sign to access the variables within the data frame:
```{r}
head(BlackWhite_results$X.2) #first six
tail(BlackWhite_results$X.2) #last six
identical(BlackWhite_results$X.1, BlackWhite_results$X.2) #are these variables the same?
```
it looks like they just count the rows. Let's remove them from the dataset:

Data manipulation is best done using functions from the package "dplyr"

install.packages("dplyr") #if you haven't installed this package before, you need to do it now!
```{r}
library(dplyr) #load this package
```
notice the output in the console. These pink notes are called "messages"
These messages are telling you that the functions "filter" and "lag" from dplyr are named the same thing
as functions from the pre-loaded stats package. So now when you call the function filter() or lag()
you will use the dplyr version of those functions. It also has functions that are the same names as 
four functions from the base R package.

Use the select() function from dplyr to select the rows you want to keep in the dataset. 
Here we use subtractive select with the negative sign to remove the rows.

```{r}
cleaned.data <- select(BlackWhite_results, -X, -X.1, -X.2)
```
notice how the number of variables has changed in the environment tab vs. the original data

now let's erase cleaned.data using the remove function, rm():
```{r}
rm(cleaned.data)
```

and recreate it using the piping operator:
```{r}
cleaned.data <- BlackWhite_results %>% select(-X, -X.1, -X.2)
```

We can View() cleaned.data or remind ourselves of the variables contained in the data frame:
```{r}
names(cleaned.data)
```

let's examine some of the variables a little more closely
```{r}
table(cleaned.data$sex)
table(cleaned.data$state)
table(cleaned.data$year)
table(cleaned.data$state, cleaned.data$Census_Region)
```

also useful is the unique() function:
```{r}
unique(cleaned.data$sex)
unique(cleaned.data$Census_Division)
length(unique(cleaned.data$Census_Division))
```

much less useful for continuous data!:
```{r}
unique(cleaned.data$LE_black_lcl)
```

```{r}
str(cleaned.data$state)
```
let's talk about factors 
factors have levels and levels are ordered.
by default factors are ordered alphabetically if they are words.
```{r}
levels(cleaned.data$state) #this prints a vector of the factor levels in their order
length(levels(cleaned.data$state)) #this prints the length
```

some other useful dplyr functions:
1. filter()
2. group_by(), summarise()
```{r}
Alabama.data <- cleaned.data %>% filter(state == "Alabama", sex == "Male")
```

```{r}
Alabama.data %>% summarise(overall.mean.white = mean(LE_white_mean), overall.mean.black = mean(LE_black_mean))
```
EXERCISE: extend the above statement to also compute the mean of the variable "LE_wbgap_mean"

Rather than doing the above in two separate steps you can "pipe" the steps together:

```{r}
cleaned.data %>%
  filter(state == "Alabama", sex == "Male") %>%
  summarise(overall.mean.white = mean(LE_white_mean), overall.mean.black = mean(LE_black_mean))
```

What if you wanted to compute the summary for females and for males?
Hint: we need to use the group_by() function as one of our commands, and also slightly modify the filter() command.

```{r}
#SOLUTION

```


EXERCISE: Compute the summary for females and males in every state.

```{r}
#SOLUTION
```

######################################################################################################
What we've learned so far:

IMPORTING DATA
read.csv() for importing .csv data. There is also read.table for .txt data
to read data exported from state you will need to load the foreign library and use read.dta()
to read files exported from Stata version 13 you need to load the readstata13() package and use read.dta13()

LOOKING AT THE DATASET
View(), str(), head(), tail(), table(), length(), unique()

summary() is another good function. Useful for continuous data:
```{r}
summary(Alabama.data$LE_white_mean)
```

and so is quantile()
```{r}
quantile(Alabama.data$LE_black_mean)
quantile(Alabama.data$LE_black_mean, 0.3)
```

MANIPULATE THE DATA
select(), filter(), group_by(), summarise()
Note that summarise computes a variety of summary functions -- our example focused on computing the mean.
Other common summary functions: median(), sum(), first(), last(), n(), min(), max(), sd()

Summarise the number of rows per stratum:
```{r}
cleaned.data %>% 
  group_by(sex, state) %>%
  summarise(num.per.strata = n())
```
Alternatively, we could have written:
```{r}
cleaned.data %>% 
  group_by(sex, state) %>%
  tally()
```
This is really useful if you have panel data and want to know the number of visits per patient, say.
Or, before running a multi-level model, you could check the cluster sizes for all your clusters

################################################################################################################

#SECTION 2: DATA VISUALIZATION

In R, there are "base R graphics", but most folks prefer the ggplot2 graphing library
```{r}
library(ggplot2) #remember to first install the package if you haven't done so already
```

the main function in ggplot2 is "ggplot"!

this function creates a plotting window that reflects the range of X and Y in the data.frame
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean))
```

we need to tell it how to plot the time series of the mean life expectancy for whites.
Let's start by plotting points:
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) +    
  geom_point()                                  
```

NOTE: the plus sign needs to always be at the end of the line. The following code WILL NOT add the geom_point() command:
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) + geom_point() 
```

Within the geom_point() function, you can customize the appearance of the points:
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) + 
  geom_point(col = "red", size = 2, pch = 3)
```

Would you prefer to have a line instead?
EXERCISE: how do you think you'd plot a line in ggplot? Give it a go. Hint: change geom_point to another type of "geom"

```{r}
#SOLUTION

```

EXERCISE: Titles
ggtitle("This is a title") adds a title to the ggplot. Add one!
Also use xlab() and ylab() to update the assigned x and y labels.

```{r}
#SOLUTION
```

How can we choose nice colours?
Nice online colour pickers:
http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
http://tristen.ca/hcl-picker/#/hlc/6/1/15534C/E2E062
pick a colour you like and copy its HEX value, including the "#". Then specify col = "#12345", where 12345 is the HEX value.
some nice colours: #1c9099, #7F526D

```{r}
#SOLUTION

```

It would be nice to also plot the life expectancy for females during the same time period.

EXERCISE: user dplyr functions on the cleaned.data to create and store a new dataset that 
contains the Alabama data for both males and females. Call the new dataset Alabama.both.genders:

```{r}
#SOLUTION

```

Let's plot the Alabama data using the previous code: 
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_line(col = "#7F526D", size = 2) + #ho boy, what is happening here? any guesses?
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Try:
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_point(col = "#7F526D", size = 2) + #better, but still not ideal. What would make this better?
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Option 1: Different colours for each gender:
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_point(aes(col = sex), size = 2) + 
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Option 2: Separate panels (aka "facets") for each gender:
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_point(size = 2) + 
  facet_wrap(~sex) +
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Let's use these same skills to look at a larger dataset

EXERCISE
PART 1: Use dplyr functions to create a new dataset containing data on only females. Call it data.females
You now have time series data for 40 states for black and white females.

```{r}
#SOLUTION
```

PART 2: plot the mean life expectancy for white females using geom_point or geom_line. 
Use one of the options learned in the previous step to plot a separate line for each state.
Which option do you like best?

```{r}
#SOLUTION A
```

```{r}
#SOLUTION B
```


For the previous plot, it would be really helpful if we could more easily tell which line is which
Plotly is a graphing package that adds interactive components: hover, zoom, highlighting
For many ggplots, you can pipe them to ggplotly() to easily add these interactive components:

```{r}
library(plotly) # don't forget to install first if you haven't done so already
```

here, we're saving the ggplot plot and calling it "our.plot"
```{r}
our.plot <- ggplot(data = data.females, aes(x = year, y = LE_white_mean)) + 
  geom_line(aes(col = state)) + 
  xlab("Year") + ylab("Mean life expectancy (years)")
```

then, we send the ggplot to the plotly() function ggplotly() to add hover text:
```{r}
our.plot %>% ggplotly()
```

EXERCISE: Incorporate information on Census_Divsion or Census_Region into the ggplot to make it more readable. 
Hint: Use both aes(col) and facet_wrap(~). Use one of the census variables, as well as the state variable.
Once you have a plot you like, make it interactive using ggplotly.

first, remind yourself what the census variables are named:
```{r}
names(data.females)
```

```{r}
#SOLUTION

```

######################################################################################################
What have we learned so far:
geom_point(), geom_line()
how to set colour to a fixed value, e.g., col("red") or col(#7F526D)
how to set colour according to the value of a variable, e.g. aes(col = state). The aes() wrapper is crucial!
facet_wrap() to make separate panels for each level of a variable
labels: ggtitle(), xlab(), ylab()
ggplotly() to make a ggplot interactive. Works on most ggplots!
######################################################################################################

EXERCISE 1: Start with cleaned.data() and use a dplyr function to keep only data from the years 1969 and 2013
HINT: In R "|" is the OR operator and "&" is the AND operator.
Use one of these operators inside a filter() statement to fiter these two years of data

```{r}
#SOLUTION 1:

```

Alternatively, use the %in% operator. 
```{r}
#SOLUTION 2:

```

EXERCISE 2: make a histogram of black life expectancy, with separate panels for year and sex.
You'll need to use two new functions: geom_histogram() and facet_grid(var1~var2).
HINT: Since histograms are univariate, you only need to set x in the ggplot() command.

```{r}
#SOLUTION

```

EXERCISE 3: This is kind of ugly. Change the colour. Did that do as you expected? Try also changing the fill.

```{r}
#SOLUTION 1:

```

```{r}
#SOLUTION 2:

```

EXERCISE 4: Update the colour to "white", and change the fill to be a function of sex. Remember aes()!

```{r}
#SOLUTION:

```

If you don't like the grey background, an easy way to remove it is to modify the underlying theme:  
tab through the themes and apply them to find the one you like.

```{r}
#SOLUTION:

```

################################################################################################################
