---
title: "Workshop Part 2"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Section 1: Viewing and Analyzing the Data

### Importing the Data
The code below reads in the `BlackWhite_results.csv` comma-separated file. This data structure has all the functionalities of a `data.frame`.

```{r}
BlackWhite_results <- read.csv("../Data/BlackWhite_results.csv")
```


### Examining the Data
Let's first use the viewer to look at these data. Click the data icon next to BlackWhite_results in the environment tab Or run the code, or run the cell below.
```{r}
View(BlackWhite_results)
```

These are some results from a study I did on the inequality in life expectancy between blacks and whites in the US. Let's first examine the structure of the dataset:

```{r}
str(BlackWhite_results)
```

what are the first three variables? 
Hint: Use the dollar sign to access the variables within the data frame.
```{r}

```

Are the X.1 and X.2 variables the same?
Write code in the cell to determine this.
```{r}

```


###Data Manipulation
It looks like they just count the rows. Let's remove them from the dataset:

Data manipulation is best done using functions from the package "dplyr"

Install.packages("dplyr"). If you haven't installed this package before, you need to do it now!
```{r}
library(dplyr) #load this package
```

Notice the output in the console. These pink notes are called "messages". These messages are telling you that the functions "filter" and "lag" from dplyr are named the same thing as functions from the pre-loaded stats package. So now when you call the function filter() or lag() you will use the dplyr version of those functions. It also has functions that are the same names as four functions from the base R package.

Use the select() function from dplyr to select the rows you want to keep in the dataset and assign it to the variable cleaned.data
Hint: Use subtractive select with the negative sign to remove the rows you do not want.
```{r}
cleaned.data <- ##replace this with code
```

Notice how the number of variables has changed in the environment tab vs. the original data.

Now let's erase cleaned.data using the remove function, rm():
```{r}
rm(cleaned.data)
```

And recreate it using the piping operator:
```{r}
cleaned.data <- BlackWhite_results %>% select(-X, -X.1, -X.2)
```


We can View() cleaned.data or remind ourselves of the variables contained in the data frame:
```{r}
names(cleaned.data)
```


Let's examine some of the variables a little more closely:
```{r}
table(cleaned.data$sex)
table(cleaned.data$state)
table(cleaned.data$year)
table(cleaned.data$state, cleaned.data$Census_Region)
```

Also useful is the unique() function:
```{r}
unique(cleaned.data$sex)
unique(cleaned.data$Census_Division)
length(unique(cleaned.data$Census_Division))
```


Much less useful for continuous data!:
```{r}
unique(cleaned.data$LE_black_lcl)

str(cleaned.data$state)
```

###Factors
Factors have levels and levels are ordered. By default, factors are ordered alphabetically if they are words.
```{r}
levels(cleaned.data$state)  #this prints a vector of the factor levels in their order
length(levels(cleaned.data$state))  #this prints the length
```

Some other useful dplyr functions:
1. filter()
2. group_by(), summarise()
```{r}
Alabama.data <- cleaned.data %>% filter(state == "Alabama", sex == "Male")

Alabama.data %>% summarise(overall.mean.white = mean(LE_white_mean), overall.mean.black = mean(LE_black_mean))
```

EXERCISE: Extend the above statement to also compute the mean of the variable "LE_wbgap_mean"
Hint: Rather than doing the above in two separate steps you can "pipe" the steps together. Replace the dots with your code.
```{r}
cleaned.data %>%
  ...  %>%
  ...
```


What if you wanted to compute the summary for females and for males?
Hint: we need to use the group_by() function as one of our commands, and also slightly modify the filter() command.
SOLUTION: We will fill this in together, and I will send a complete file with solutions after class.
```{r}
cleaned.data %>% 
  group_by(sex) %>%
  filter(state == "Alabama") %>%
  summarise(overall.mean.white = mean(LE_white_mean), overall.mean.black = mean(LE_black_mean))
```

EXERCISE: Compute the summary for females and males in every state.
TO BE FILLED IN DURING CLASS
```{r}
cleaned.data %>% 
  ... %>%
  summarise(overall.mean.white = ... , 
            overall.mean.black = ...)) %>% 
  print(n = nrow(.)) # add the print step to specify the number of rows to print. This tells it to print all rows of the results.
```

######################################################################################################
###What we've learned so far:

1. IMPORTING DATA
a. read.csv() for importing .csv data. There is also read.table for .txt data to read data exported from state you will need to load the foreign library and use read.data()
b. To read files exported from Stata version 13 you need to load the readstata13() package and use read.dta13()

2. LOOKING AT THE DATASET
a. View(), str(), head(), tail(), table(), length(), unique()
b. summary() is another good function. Useful for continuous data: summary(Alabama.data$LE_white_mean)
c. And so is quantile():
quantile(Alabama.data$LE_black_mean)
quantile(Alabama.data$LE_black_mean, 0.3)

3. MANIPULATE THE DATA
a. select(), filter(), group_by(), summarise()
Note that summarise computes a variety of summary functions -- our example focused on computing the mean.
Other common summary functions: median(), sum(), first(), last(), n(), min(), max(), sd()

4. Summarise the number of rows per stratum:
cleaned.data %>% 
  group_by(sex, state) %>%
  summarise(num.per.strata = n())
Alternatively, we could have written:
cleaned.data %>% 
  group_by(sex, state) %>%
  tally()
This is really useful if you have panel data and want to know the number of visits per patient, say.
Or, before running a multi-level model, you could check the cluster sizes for all your clusters

################################################################################################################

###SECTION 2: DATA VISUALIZATION
In R, there are "base R graphics", but most folks prefer the ggplot2 graphing library
```{r}
library(ggplot2) #remember to first install the package if you haven't done so already
```

The main function in ggplot2 is "ggplot"!  This function creates a plotting window that reflects the range of X and Y in the data.frame
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean))
```

We need to tell it how to plot the time series of the mean life expectancy for whites. Let's start by plotting points:
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) +    
  geom_point()  
```
                                
NOTE: the plus sign needs to always be at the end of the line. The following code WILL NOT add the geom_point() command:
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean))
+ geom_point() 
```


Within the geom_point() function, you can customize the appearance of the points:
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) + 
  geom_point(col = "red", size = 2, pch = 3)
```


Would you prefer to have a line instead?
EXERCISE: How do you think you'd plot a line in ggplot? Give it a go. 
Hint: change geom_point to another type of "geom"
```{r}

```

EXERCISE: Titles
ggtitle("This is a title") adds a title to the ggplot. Add one!
Also use xlab() and ylab() to update the assigned x and y labels.
```{r}

```


How can we choose nice colours?
Nice online colour pickers:
http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
http://tristen.ca/hcl-picker/#/hlc/6/1/15534C/E2E062
Pick a colour you like and copy its HEX value, including the "#". Then specify col = "#12345", where 12345 is the HEX value. 
Some nice colours: #1c9099, #7F526D
Replace the dots in the code with your favorite colors. Pick colors that are easy on the eyes! (Note: make sure to wrap the colors around
with quotes, your code will not run!)
```{r}
ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) + 
  geom_line(col = ... , size = 2) +
  xlab("Year") + ylab("Mean life expectancy (years)") +
  ggtitle("Mean life expectancy for white males in Alabama, 1968--2013")

ggplot(data = Alabama.data, aes(x = year, y = LE_white_mean)) + 
  geom_line(col = ... , size = 2) +
  xlab("Year") + ylab("Mean life expectancy (years)") +
  ggtitle("Mean life expectancy for white males in Alabama, 1968--2013")
```


It would be nice to also plot the life expectancy for females during the same time period.
EXERCISE: user dplyr functions on the cleaned.data to create and store a new dataset that contains the Alabama data for both males and females. Call the new dataset Alabama.both.genders:
```{r}
Alabama.both.genders <- cleaned.data %>% ...
```

Let's plot the Alabama data using the previous code: 
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_line(col = "#7F526D", size = 2) + ##oh boy, what is happening here? any guesses?
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Try:
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_point(col = "#7F526D", size = 2) + #better, but still not ideal. What would make this better?
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Option 1: Different colours for each gender:
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_point(aes(col = sex), size = 2) + 
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Option 2: Separate panels (aka "facets") for each gender:
```{r}
ggplot(data = Alabama.both.genders, aes(x = year, y = LE_white_mean)) + 
  geom_point(size = 2) + 
  facet_wrap(~sex) +
  xlab("Year") + ylab("Mean life expectancy (years)") 
```

Let's use these same skills to look at a larger dataset

EXERCISE
PART 1: Use dplyr functions to create a new dataset containing data on only females. Call it data.females
You now have time series data for 40 states for black and white females.
```{r}
data.females <- ...
```

PART 2: Plot the mean life expectancy for white females using geom_point or geom_line. 
Use one of the options learned in the previous step to plot a separate line for each state.
Which option do you like best?
```{r}

```

For the previous plot, it would be really helpful if we could more easily tell which line is which. 
Plotly is a graphing package that adds interactive components: hover, zoom, highlighting. For many ggplots, you can pipe them to ggplotly() to easily add these interactive components:
```{r}
library(plotly) # don't forget to install first if you haven't done so already
```

Here, we're saving the ggplot plot and calling it "our.plot"
```{r}
our.plot <- ggplot(data = data.females, aes(x = year, y = LE_white_mean)) + 
  geom_line(aes(col = state)) + 
  xlab("Year") + ylab("Mean life expectancy (years)")

our.plot #this prints the static version
```

Then, we send the ggplot to the plotly() function ggplotly() to add hover text:
```{r}
our.plot %>% ggplotly()
```


EXERCISE: Incorporate information on Census_Divsion or Census_Region into the ggplot to make it more readable. 
Hint: Use both aes(col) and facet_wrap(~). Use one of the census variables, as well as the state variable.
Once you have a plot you like, make it interactive using ggplotly.


First, remind yourself what the census variables are named:
```{r}
names(data.females)
```

```{r}
##Type solution here


```


######################################################################################################
What have we learned so far:
a. geom_point(), geom_line()
b. how to set colour to a fixed value, e.g., col("red") or col(#7F526D)
c. how to set colour according to the value of a variable, e.g. aes(col = state). The aes() wrapper is crucial!
d. facet_wrap() to make separate panels for each level of a variable
e. labels: ggtitle(), xlab(), ylab()
f. ggplotly() to make a ggplot interactive. Works on most ggplots!
######################################################################################################

EXERCISE 1: Start with cleaned.data() and use a dplyr function to keep only data from the years 1969 and 2013
HINT: In R "|" is the OR operator and "&" is the AND operator.
Use one of these operators inside a filter() statement to fiter these two years of data
```{r}

```

Alternatively, use the %in% operator. 
```{r}

```


EXERCISE 2: Make a histogram of black life expectancy, with separate panels for year and sex. Call your histogram two.years
You'll need to use two new functions: geom_histogram() and facet_grid(var1~var2).
HINT: Since histograms are univariate, you only need to set x in the ggplot() command.
```{r}

```

EXERCISE 3: This is kind of ugly. Change the colour. Did that do as you expected? Try also changing the fill.
```{r}

```


EXERCISE 4: Update the colour to "white", and change the fill to be a function of sex. Remember aes()!
```{r}

```

If you don't like the grey background, an easy way to remove it is to modify the underlying theme:  
Tab through the themes and apply them to find the one you like.
```{r}

```


################################################################################################################

###SECTION 3: ANOTHER DATASET, ANOTHER DATA VISUALIZATION
I previously saved the R data.frame "CS.data" inside R. We can reload it using the load() command:
```{r}
CS.data <- read.csv("../Data/Cesarean.csv")
```

Notice that CS.data has been added to the Global Environment.
Check out CS.data using some of the commands we learned earlier. There is a variable called "X2006". This variable is the gross domestic product in 2006 for the corresponding country.
```{r}



```


Let's use the dplyr function rename() to give this variable a better name.
```{r}
example.data <- example.data %>% rename(new.name = old.name)
```

```{r}
CS.data <- CS.data %>% rename(gdp.2006 = X2006)
```


CS_Rate2 is the cesarean delivery rate as a proportion < 1. For graphing purposes, it will be nice to have this variable have an upper bound of 100.
Use the dplyr function mutate() to create a new variable based on CS_Rate2 that is multiplied by 100.
```{r}
example.data <- example.data %>% mutate(mutated.var = function(existing.var))

CS.data <- CS.data %>% mutate(...)
```

Drop two of the country variables and keep the third. Use View() to look at the values and choose the best variable to keep.
Also drop the old CS_Rate2 variable
```{r}
CS.data <- CS.data %>% select(-Country.y, -Country.x, -CS_Rate2)
```

EXERCISE: Visually assess the relationship between GDP and cesarean delivery rate using a scatter plot (i.e., geom_point).
Start simple and then keep enhancing the ggplot until it is awesome.
Remember: col(), fill(), and size(). Also, remember the themes.
Experiment with geom_text(). Type ?geom_text into the console to open a help window and learn about it or google it!

The very basics
```{r}
ggplot(data = CS.data, aes(x = ..., y = ...)) + 
  geom_point()
```

There is a lot of data smushed near 0, especially on the x-axis -- could a transformation help?
```{r}
ggplot(data = ..., aes(x = ... , y = ...)) + 
  geom_point(aes(...)) +
  ... ##replace these dots with aesthetic
```

where should size of the points come in?
```{r}
ggplot(data = CS.data, aes(...)) + 
  geom_point(aes(...)) +
  ...
```

I want the points to all be BIGGER! You can adjust the max point size using scale_size_area(max_size = ##)
```{r}
ggplot(data = CS.data, aes(...)) + 
  geom_point(...) +
  ... + 
  ...
```

Oh boy, the legend is massive -- let's remove it. And some points are overlapping -- you can make the points semi-transparent by setting the alpha parameter to a value between 0 and 1:
```{r}
ggplot(data = CS.data, aes(...) + 
  geom_point(..., alpha = ...)) +
  ... + 
  scale_size_area(... , guide = "none")
```

which country is which?
```{r}
ggplot(data = CS.data, aes(x ...Region, size = ..., alpha = ...)) +
  ... + 
  scale_size_area(...) +
  geom_text(...)

```

Too many labels. Let's use an ifelse() statement to selectively display labels for a certain set of countries
```{r}
ggplot(data = CS.data, aes(...) + 
  geom_point(...) +
  ... +
  scale_size_area(..., guide = "none") +
  geom_text(aes(label = ifelse(... , ...)

ggplot(data = CS.data, aes(...) + 
  geom_point(...) +
  ... +
  scale_size_area(..., guide = "none") +
  geom_text(aes(label = ifelse(... , as.character(Country_Name), "")))
```


Some more easy style changes
```{r}
ggplot(data = CS.data, aes(x = gdp.2006, y = CS_Rate)) + 
  geom_point(aes(col = Region, size = Births_Per_1000, alpha = 0.5)) +
  scale_x_log10() +
  scale_size_area(max_size = 50, guide = "none") +
  scale_alpha_continuous(guide = "none") +
  geom_text(aes(label = ifelse(Births_Per_1000 > 1000, as.character(Country_Name), ""))) +
  xlab("GDP") + ylab("Cesarean delivery rate (%)") + theme_minimal()

cs.plot <- ggplot(data = CS.data, aes(x = gdp.2006, y = CS_Rate)) + 
  geom_point(aes(fill = Region, size = Births_Per_1000, alpha = 0.5), col = "black", shape = 21) +
  scale_x_log10(breaks = c(1, 100, 1000, 10000), labels = c(1, 100, 1000, 10000)) +
  scale_size_area(max_size = 50, guide = "none") +
  scale_alpha_continuous(guide = "none") +
  geom_text(aes(label = ifelse(Births_Per_1000 > 700, as.character(Country_Name), ""))) +
  xlab("GDP") + ylab("Cesarean delivery rate (%)") + theme_minimal() 
  
cs.plot %>% ggplotly()
```


######################################################################################################
What else did we learn:
a. dplyr functions rename() to change variable names, and mutate() to add new variables based on existing variables
b. alpha() to control transparency, shape() to change shape of geom_point()
c. scale_x_log10() to use a log-transformed x-axis
d. geom_text() to add labels to the plot
e. ifelse() operator. If true then do: ...., else, do: ....

Really useful resources for learning ggplot2 and dplyr:
ggplot2 cheatsheet: https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
dplyr cheatsheet: https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
######################################################################################################
